\chapter{Repeats}

To replace a bar line with a left, right, or left-right repeat, use one of the
commands \keyindex{leftrepeat}, \keyindex{rightrepeat} or
\keyindex{leftrightrepeat} in place of \verb|\bar|. If a \verb|\leftrepeat|
happens to come at the end of a system, it will automatically be moved to the
start of the next system. If a \verb|\leftrightrepeat| happens to come at the
end of a system, \musixtex\ will automatically post a right repeat at the end
of the system and a left repeat at the beginning of the next.

For example,

\begin{music}\nostartrule
\startextract
\NOTes\ha g\en
\leftrepeat
\NOTes\ha h\en
\leftrightrepeat
\NOTes\ha i\en
\rightrepeat
\NOTEs\wh j\en
\zendextract
\end{music}
 \noindent has been coded as:
\begin{quote}\begin{verbatim}
\NOTes\ha g\en
\leftrepeat
\NOTes\ha h\en
\leftrightrepeat
\NOTes\ha i\en
\rightrepeat
\NOTEs\wh j\en
\end{verbatim}\end{quote}

To insert a right-repeat at a forced line break or at the end of a piece, use
\verb|\setrightrepeat| \textit{before}\ \verb|\alaligne| or \verb|\endpiece|.
In fact it is possible to use \keyindex{setleftrepeat},
\keyindex{setrightrepeat} or \keyindex{setleftrightrepeat} before any
\keyindex{bar}, \keyindex{stoppiece} or \keyindex{changecontext}; but
be aware that whereas \verb|\setrightrepeat| behaves properly if the bar is at the
end of a system, \verb|\setleftrepeat| and \verb|\setleftrightrepeat| do not,
placing
the left-repeat symbol at the end of the system rather
than at the beginning of the next system; see the 
\verb|\zleftrepeat| \verb|\zzleftrepeat| \verb|\zzleftrightrepeat| 
and \verb|\zzleftrightrepeat| commands described below.


The following commands generate repeat symbols as usual, but
are assumed to be in the \emph{middle} of a bar; that is,
the bar counter is not incremented:
\begin{itemize}
\item[] \keyindex{zrightrepeat}
\item[] \keyindex{zleftrepeat}
\item[] \keyindex{zleftrightrepeat}
\end{itemize}
For example, to insert a left-repeat at a forced line break or at the
beginning of a piece, simply use \verb|\zleftrepeat| immediately \textit{after}
\verb|\startpiece| or \verb|\alaligne|. To insert a left-right repeat at a
forced line break, use \verb|\setrightrepeat\alaligne\zleftrepeat|. 

The following commands are similar but force the mid-bar break to the
end of a system:
\begin{itemize}
\item[] \keyindex{zzrightrepeat}
\item[] \keyindex{zzleftrepeat}
\item[] \keyindex{zzleftrightrepeat}
\end{itemize}
For
\verb|\zzleftrepeat| and \verb|\zzleftrightrepeat|, the left-repeats occur
at the beginning of the following system. The bar numbers are
correctly adjusted, using \keyindex{writezbarno} for the continuation
if \keyindex{systemnumbers} is in use. 

As an example, here is
a \emph{bour\'ee} by G.-F.~Handel coded using \verb|\zleftrepeat| at the beginning
of the \textbf{A} section and \verb|\zzleftrightrepeat| 
between the \textbf{A} and the \textbf{B} sections:

\begin{music}
\smallmusicsize
\generalmeter{\allabreve}
\generalsignature{1}
\startbarno=0
\nobarnumbers
\nostartrule
\startpiece\addspace\afterruleskip
\systemnumbers
\zleftrepeat%
\znotes\ccn{10}{\boxit{\bf A}}\en
\NOtes\qa d\en%
\bar%1
\NOtes\qa{g}\qa{i}\qa{h}\qa{g}\en%
\bar%2
\NOTes\ha{k}\ha{i}\en%
\bar%3
\NOtes\qa{l}\qa{k}\qa{j}\qa{i}\en%
\bar%4
\NOtes\qa j\en
\NOTes\ha i\en%
\zzleftrightrepeat%
\znotes\ccn{10}{\boxit{\bf B}}\en
\NOtes\qa k\en%
\bar%5
\NOtes\qa{l}\qa{j}\qa{h}\en
\Notes\ibl3i2\qb3i\tql3j\en%
\bar%6
\NOtes\qa{k}\qa{i}\qa{g}\en
\Notes\ibu3h2\qb3h\tqu3i\en%
\bar%7
\NOtes\qa{j}\qa{i}\qa{h}\qa{g}\en%
\bar%8
\Notes\ibu3h1\qb3{h}\qb3{g}\qb3{h}\tqu3i\en
\NOtes\qa h\en
\Notes\ibl3i2\qb3i\tql3j\en%
\bar%9
\NOtes\qa k\en
\Notes\ibl3i2\qb3i\tql3j\en
\NOtes\qa k\en
\Notes\ibl3j{-2}\qb3j\tql3i\en%
\bar%10
\NOtes\qa j\en
\Notes\ibu3h2\qb3h\tqu3i\en
\NOtes\qa j\en
\Notes\ibu3i{-2}\qb3i\tqu3h\en%
\bar%11
\NOtes\qa{i}\qa{g}\en
\NOtesp\tr{11}\qa{.h}\en
\Notes\ca g\en%
\bar%12
\NOTesp\ha{.g}\en%
\setrightrepeat%
\endpiece%
\end{music}

\noindent No explicit adjustment of \verb|\barno| was necessary.  The right-repeat
at the end of the \textbf{B}~section was produced by
\verb|\setrightrepeat\endpiece|.



 \section{First and second endings (Voltas)}\index{volta}

All volta commands must be entered right before the bar line command
(or repeat, etc.) where they are to take effect. There are three
commands that suffice to set all voltas. To start one, use
\keyindex{Setvolta}\verb|{|\textit{text}\verb|}|; to terminate it
with or without a vertical line, use \keyindex{endvolta} or
\keyindex{endvoltabox} respectively. The text by default will be followed
by a period. There are also various alternate
commands (e.g., \verb|\setendvoltabox| is equivalent to \verb|\endvoltabox|).
Some such alternate forms are used in the following example, but the first three
mentioned above are all that are required:

\medskip \begin{music}
 \parindent0pt
 \nostartrule
 \startpiece
 \addspace\afterruleskip
 \NOTEs\wh a\en\bar
 \NOTEs\wh b\en\setvoltabox{1.-3}\bar
 \NOTEs\wh c\en\setvolta4\setendvolta\rightrepeat
 \NOTEs\wh d\en\doublebar
 \NOTEs\wh e\en\bar
 \NOTEs\wh f\en\leftrepeat
 \NOTEs\wh g\en\bar
 \NOTEs\wh h\en\Setvolta1\bar
 \NOTEs\wh i\en\bar
 \NOTEs\wh j\en\Setvolta2\setendvoltabox\rightrepeat
 \NOTEs\wh i\en\bar
 \NOTEs\wh h\en\setendvoltabox
 \Endpiece
 \end{music}
 \noindent This was coded as
 \begin{quote}\begin{verbatim}
 \startpiece \addspace\afterruleskip
 \NOTEs\wh a\en\bar
 \NOTEs\wh b\en\setvoltabox{1.-3}\bar
 \NOTEs\wh c\en\setvolta4\setendvolta\rightrepeat
 \NOTEs\wh d\en\doublebar
 \NOTEs\wh e\en\bar
 \NOTEs\wh f\en\leftrepeat
 \NOTEs\wh g\en\bar
 \NOTEs\wh h\en\Setvolta1\bar
 \NOTEs\wh i\en\bar
 \NOTEs\wh j\en\Setvolta2\setendvoltabox\rightrepeat
 \NOTEs\wh i\en\bar
 \NOTEs\wh h\en\setendvoltabox
 \Endpiece
\end{verbatim}\end{quote}
 \zkeyindex{leftrepeat}\zkeyindex{rightrepeat}

If the volta only spans one measure and ends without a vertical segment,
it can be specified simply by saying
\keyindex{setvolta}\verb|{|\textit{text}\verb|}| before the bar line command
that starts it, and it will automatically terminate at the
second bar line command:

\medskip \begin{music}
 \parindent0pt \nostartrule\startpiece \addspace\afterruleskip
 \NOTEs\wh a\en\bar
 \NOTEs\wh b\en\setvolta{1.-3}\bar \NOTEs\wh c\en\setvolta4\rightrepeat
 \NOTEs\wh d\en\bar
 \NOTEs\wh e\en\Endpiece
\end{music}
\noindent which was coded as:
 \begin{quote}\begin{verbatim}
 \parindent0pt \startpiece \addspace\afterruleskip
 \NOTEs\wh a\en\bar
 \NOTEs\wh b\en\setvolta{1.-3}\bar \NOTEs\wh c\en\setvolta4\rightrepeat
 \NOTEs\wh d\en\bar
 \NOTEs\wh e\en\Endpiece
\end{verbatim}\end{quote}

The height above the top staff line of the horizontal line in a volta symbol
is determined by the token \keyindex{raisevolta} which is \verb|4\internote| by
default. You can change this to any desired dimension.

The period after the text can be
removed by saying \verb|\def|\keyindex{voltadot}\verb|{}|
and restored by \verb|\def|\keyindex{voltadot}\verb|{.}|\ .

 \section{Special symbols for repeating long sections}

Four special symbols and corresponding macros are available, namely
\keyindex{coda}~$p$, \keyindex{Coda}~$p$,
and \keyindex{segno}~$p$, where $p$ specifies the pitch; and \keyindex{Segno}
with no argument. Their behavior is illustrated in this example:

\begin{music}
\startextract
\NOtes\segno n\en
\bar
\NOtes\coda n\en
\NOtes\Segno\en
\bar
\NOtes\Coda n\en
\endextract
\end{music}
\noindent which has been coded:
\begin{quote}\begin{verbatim}
\NOtes\segno m\en\bar
\NOtes\coda m\en
\NOtes\Segno\en\bar
\NOtes\Coda m\en
\end{verbatim}\end{quote}

\section{Repeating a single bar}
 The special symbol for a single-bar repeat is generated by \keyindex{duevolte}, as shown
in the following example:

\begin{music}
\generalmeter\meterC
\setclef1\bass\setstaffs1{2}
\startextract
\NOtes|\qa{cegj}\en
\def\atnextbar{\znotes\centerpause|\en}\bar
\NOtes\qa{cdef}|\en
\def\atnextbar{\znotes|\centerbar{\duevolte}\en}%
\endextract
\end{music}
\noindent whose coding is:
\begin{quote}\begin{verbatim}
\generalmeter\meterC
\setclef1\bass\setstaffs1{2}
\startextract
\NOtes|\qa{cegj}\en
\def\atnextbar{\znotes\centerpause|\en}\bar
\NOtes\qa{cdef}|\en
\def\atnextbar{\znotes|\centerbar{\duevolte}\en}%
\endextract
\end{verbatim}\end{quote}
