\chapter{Rests}
 \section{Ordinary rests}
 A separate macro is defined for each kind of ordinary rest. They cause
a space after the symbol, just like spacing note commands, but they have
no parameters. A whole rest is coded as
\keyindex{pause}, 
half rest \keyindex{hpause} or \keyindex{hp}, 
quarter rest \keyindex{qp} or \keyindex{soupir},
eighth rest \keyindex{ds}, sixteenth rest \keyindex{qs},
$32$nd~rest \keyindex{hs}, and $64$th rest \keyindex{qqs}.
Dotted rests may be obtained by 
using 
\keyindex{hpausep} or \keyindex{hpp}, \keyindex{qpp}, \keyindex{dsp}, \keyindex{qsp},
\keyindex{hsp} and \keyindex{qqsp}.

 Longer rests, normally interpreted as lasting
two or four bars respectively, can be coded as \keyindex{PAuse}
and \keyindex{PAUSe}, which yield:

\begin{music}\nostartrule
\generalmeter{\meterfrac44}
\startextract
\def\atnextbar{\znotes\centerbar\PAuse\en}%
\NOTEs\en\bar
\def\atnextbar{\znotes\centerbar\PAUSe\en}%
\NOTEs\en
\endextract
\end{music}
 %\check


 \section{Raising rests}\index{raising rests}
All the
previous rests except \keyindex{pausep} and
\keyindex{hpausep} are \ital{hboxes}, which means that
they can be vertically offset if needed using the
standard \TeX\ command \keyindex{raise}. For example:

 \begin{quote}
 \begin{verbatim}
 \raise 2\Interligne\qp
 \raise 3mm\qqs
 \end{verbatim}
 \end{quote}

\noindent where \keyindex{Interligne} is the distance from one staff line to the
next.

In addition, two macros are available to put a whole or
half rest above or below the staff. The ordinary \verb|\pause| or
\verb|\hpause| cannot be used outside the staff because a short horizontal line
must be added to distinguish between the whole and the half rest. The commands,
which are non-spacing\footnote{Editor's note: The reason for having defined these
as non-spacing is not obvious}, are
 \begin{itemize}\setlength{\itemsep}{0ex}
 \item \keyindex{liftpause}~$n$ to get a
  \hbox to10pt{\liftpause{-2}\hss}
  raised from original position by $n$ staff line intervals,
 \item \keyindex{lifthpause}~$n$ to get
  \hbox to10pt{\lifthpause{-1}\hss} raised the same way.
 \item \keyindex{liftpausep}~$n$ to get a
  \hbox to10pt{\liftpausep{-2}\hss}
  raised from original position by $n$ staff line intervals,
 \item \keyindex{lifthpausep}~$n$ to get
  \hbox to10pt{\lifthpausep{-1}\hss} raised the same way.
 \end{itemize}
 %\check

 \section{Bar centered rests}\label{barcentered}
Sometimes it is necessary to place a rest (or any other symbol) exactly in the middle
of a bar. This can be done with combinations of the commands
\keyindex{atnextbar},
\keyindex{centerbar},
\keyindex{centerPAUSe},
\keyindex{centerPAuse},
\keyindex{centerpause},
\keyindex{centerhpause},
as demonstrated in the following example:

\begin{music}
\generalmeter\meterC
\setclef1\bass
\setstaffs1{2}
\parindent0pt
\startpiece\addspace\afterruleskip
\NOtes|\qa{cegj}\en
\def\atnextbar{\znotes\centerpause\en}\bar
\NOtes|\qa{jgec}\en
\def\atnextbar{\znotes\centerpause\en}\bar
\Notes\ca{`jihgfedc}\en
\def\atnextbar{\znotes|\centerpause\en}\bar
\NOTes\ha{Nc}\en
\def\atnextbar{\znotes|\centerpause\en}\bar
\addspace{10\elemskip}%
\def\atnextbar{\znotes\centerbar\duevolte|\centerbar\duevolte\en}\endpiece
\end{music}
\noindent with the coding
\begin{verbatim}
\generalmeter\meterC
\setclef1\bass
\setstaffs1{2}
\parindent0pt
\startpiece\addspace\afterruleskip
\NOtes|\qa{cegj}\en
\def\atnextbar{\znotes\centerpause\en}\bar
\NOtes|\qa{jgec}\en
\def\atnextbar{\znotes\centerpause\en}\bar
\Notes\ca{`jihgfedc}\en
\def\atnextbar{\znotes|\centerpause\en}\bar
\NOTes\ha{Nc}\en
\def\atnextbar{\znotes|\centerpause\en}\bar
\addspace{10\elemskip}%
\def\atnextbar{\znotes\centerbar{\duevolte}|\centerbar{\duevolte}\en}\endpiece
\end{verbatim}
