\chapter{Font-based Slurs, Ties and Hairpins}
Two fundamentally different implementations of slurs, ties, and hairpins are
available. First, there are the
original \ital{font-based} versions. These are constructed with traditional
\TeX\ font characters that were created with \MF\ and stored in \TeX\ font files.
Second, it is now possible
to generate these shapes directly with Postscript, dispensing altogether with
the font characters\footnote{Please do not be confused by the availability of
Postscript versions of the font-based slur fonts (along with all other \musixtex\ fonts).
Once installed in a \TeX\ system,
their function and use are 100\% transparently identical with bitmapped versions
of the slur fonts. On the other
hand, Postscript slurs are functionally distinct from font-based slurs, and only
 share some of the same syntax.}. We shall first describe the font-based versions,
then Type~K Postscript slurs, which are one of two available Postscript slur
options\footnote{An
alternate approach to Postscript slurs, called {\it Type~M} after its developer
Hiroaki {\sc Morimoto}, is available from the
\href{http://icking-music-archive.org/software/indexmt6.html}
{\underline{Icking Music Archive}}.}. If you plan to use Type~K Postscript slurs,
you may skip directly to Chapter~\ref{PostscriptSlurs}.

Font-based slurs and ties provided by \musixtex\ can be divided into two categories:
\label{simpleslur}
\begin{itemize}\setlength{\itemsep}{0ex}
   \item Those where the complete slur symbol is composed of a single
   character from one of the slur fonts, and
   \item those where the slur symbol is composed of three distinct
   characters, to form the beginning, middle and end of the slur.
\end{itemize}
The former are called \ital{simple slurs} and the latter,
\ital{compound slurs}. In many cases the distinction between the two is
invisible to the user, in that many of the macros described below will
automatically select between the two types. However, there are other macros
that allow simple slurs to be forced.

The next few sections describe the usual method of slur coding, where the
choice between simple or compound slurs is made automatically.
In this case, slurs are initiated and terminated by separate macros,
similar to beams.

 \section{Font-based slur initiation}
A slur must be initiated \ital{before}
the spacing note on which the slur begins, and terminated \ital{before} the
note on which it ends. 

The basic slur initiation macro is
\keyindex{isluru}\enpee, which initiates an upper slur, with reference
number $n$, beginning on a note at pitch $p$. The starting point of the slur is
centered above a virtual quarter note head at pitch~$p$\footnote{The slur will start in the same place regardless of whether there is
{\it actually} a note at pitch $p$.}.
Similarly, \keyindex{islurd}\enpee\ initiates a lower slur.
These slurs are terminated by coding
\keyindex{tslur}\enpee\ where $n$ is the reference number and $p$ is the
termination pitch. 

As with beams, the reference number $n$ by default can take values from $0$ to $5$, or
up to $8$ or $11$ respectively if \ttxem{musixadd.tex} or \ttxem{musixmad.tex} is included.
You can also specify the maximum number 
directly with \keyindex{setmaxslurs}\verb|{|$m$\verb|}| where 
$7<m\leq 100$%
\footnote{This may require e-\TeX.}; 
the reference number $n$ will be in the range between $0$ and $m-1$.
\label{musixmad_setmaxslurs}

To illustrate with an elementary example, the following passage

\begin{music}\nostartrule
\startextract
\NOtes\islurd0g\qu g\tslur0{'c}\qu c\en
\Notes\isluru0{'e}\ibl0e{-2}\qb0{edc}\tslur0b\tqb0b\en
\bar
\NOtes\islurd0{'a}\qu a\tslur0{`f}\qu f\en
\NOTes\hu g\en
\zendextract
\end{music}

\noindent was coded as:
\begin{quote}\begin{verbatim}
\NOtes\islurd0g\qu g\tslur0{'c}\qu c\en
\Notes\isluru0{'e}\ibl0e{-2}\qb0{edc}\tslur0b\tqb0b\en
\bar
\NOtes\islurd0{'a}\qu a\tslur0{`f}\qu f\en
\NOTes\hu g\en
\end{verbatim}\end{quote}

Other macros are provided to change the horizontal starting and ending points of the slur
in relation to the initial and final notes. Thus, \keyindex{issluru}\enpee\
initiates a ``short'' upper slur suitable for linking notes involved in chords.
The
starting point is shifted to the right, and is vertically aligned with the
center of a virtual quarter note head at pitch~$p$. If a lower short slur
is wanted, one should use \keyindex{isslurd}\enpee.

Sometimes, busy scores call for slurs which are vertically aligned with the
ends of note stems rather than note heads. These ``beam'' slurs---so called
because the slur is written at usual beam height---are provided by the
macros \keyindex{ibsluru}\enpee\ and \keyindex{ibslurd}\enpee. These macros
initiate slurs raised or lowered by the current stem height to accommodate
stems or beams above or below.
 %\check

 \section{Font-based slur termination}
Font-based slurs that are not forced to be simple must be terminated by an explicit
command right before the last note under the slur. There are termination
commands analogous to each of the initiation commands already presented. They
are summarized in the following table:
\begin{center}
  \begin{tabular}{ll}
    Initiation                             &   Termination \\
    \hline
    \keyindex{isluru},  \keyindex{islurd}  & \keyindex{tslur}  \\
    \keyindex{issluru}, \keyindex{isslurd} & \keyindex{tsslur} \\
    \keyindex{ibsluru}                     & \keyindex{tbsluru}\\
    \keyindex{ibslurd}                     & \keyindex{tbslurd}\\
    \hline
  \end{tabular} \end{center}

\noindent All of these command have two parameters, $n$ and $p$.

These specific termination macros are not restricted to
being used with their initiation counterpart. A slur started
in one sense can be terminated in another. For example, a
slur beginning as a ``beam'' slur may be terminated as a
normal slur, or \emph{vice versa}. These would be achieved using macro pairs
\verb|\ibslur...\tslur...| or \verb|\islur...\tbslur...|, respectively; however, it may be necessary
to adjust the vertical position of the slur termination. For example, 

\begin{music}
\startextract
\Notes\ibsluru1g\qa g\ha l\sk\tslur1e\zq g\zq i\qu k\en
\Notes\islurd1h\qu h\ha f\sk\tbslurd1c\zq h\zq j\ql l\en
\endextract
\end{music}\noindent
has been coded as follows:
\begin{quote}
\begin{verbatim}
\startextract
\Notes\ibsluru1g\qa g\ha l\sk\tslur1e\zq g\zq i\qu k\en
\Notes\islurd1h\qu h\ha f\sk\tbslurd1c\zq h\zq j\ql l\en
\endextract
\end{verbatim}
\end{quote}
 \section{Font-based ties} Font-based ties will have the same shapes as
ordinary font-based slurs of the same length between notes of equal pitch,
but there are two important distinctions:
(1) There cannot be any pitch difference between start and end, and (2) the
positions of both the beginning and end of a tie relative to the note heads
are slightly different from those of an ordinary slur\footnote{Editor's note: In
fact, it appears that the default positioning of the ends of ties is exactly
the same as that of short slurs.}.
Upper ties are initiated by \keyindex{itieu}\enpee, which starts an upper tie of
reference number $n$ at pitch $p$.
Lower ties are initiated by \keyindex{itied}\enpee, which starts an lower tie of
reference number $n$ at pitch $p$. The starting position of the tie is the same
as \verb|\issluru| and
\verb|\isslurd| respectively. The tie is terminated by coding
\keyindex{ttie}\onen. Note that no pitch parameter is required.

There are also \ital{short ties}, which bear the same relation to ordinary
ties as short slurs to ordinary slurs. They are intended to be used between chords.
They are initiated with
\keyindex{itenu}\enpee~or \keyindex{itenl}\enpee\footnote{Editor's note: It is
not clear why this command uses ``{\tt l}'' when all other similar ones use
``{\tt d}''.}, and terminated with \keyindex{tten}\onen.

The following example illustrates the differences in positioning of the various
slur and tie options:

\begin{music}\nostartrule
\startextract
\NOTes\islurd0g\qu g\tslur0g\qu g\isslurd0g\qu g\tsslur0g\qu g%
\ibsluru0g\qu g\tbsluru0g\qu g\itied0g\qu g\ttie0\qu g%
\itenl0g\qu g\tten0\qu g\en
\zendextract
\end{music}
\noindent It was coded as
\begin{quote}\begin{verbatim}
\NOTes\islurd0g\qu g\tslur0g\qu g\isslurd0g\qu g\tsslur0g\qu g%
\ibsluru0g\qu g\tbsluru0g\qu g\itied0g\qu g\ttie0\qu g%
\itenl0g\qu g\tten0\qu g\en
\end{verbatim}\end{quote}

\noindent Here are some more general examples of font-based slurs and ties
discussed so far:

\begin{music}\nostartrule
\startextract
\NOTes\isluru0g\hl g\tslur0h\hl h\en
\NOTes\islurd0c\issluru1g\zh{ce}\hu g\tslur0d\tsslur1h\zh{df}\hu h\en
\NOTes\ibsluru0g\islurd1g\hu g\tubslur0h\hu h\en
\NOTes\itieu0k\hl k\ttie0\tdbslur1f\hl k\en
\zendextract
\end{music}
\noindent This was coded as:
\begin{quote}\begin{verbatim}
\NOTes\isluru0g\hl g\tslur0h\hl h\en
\NOTes\islurd0c\issluru1g\zh{ce}\hu g\tslur0d\tsslur1h\zh{df}\hu h\en
\NOTes\ibsluru0g\islurd1g\hu g\tubslur0h\hu h\en
\NOTes\itieu0k\hl k\ttie0\tdbslur1f\hl k\en
\end{verbatim}\end{quote}

\section{Dotted slurs}

Any font-based slur may be made dotted by specifying \keyindex{dotted}
just before it is initiated:

\begin{music}\nostartrule
\startextract
\NOtes\dotted\islurd0g\qu g\tslur0{'c}\qu c\en
\Notes\dotted\isluru0{'e}\ibl0e{-2}\qb0{edc}\tslur0b\tqb0b\en\bar
\NOtes\dotted\slur{'a}{`f}d1\qu{'a`f}\en
\NOTes\hu g\en
\zendextract
\end{music}

\noindent This was coded as:
\begin{quote}\begin{verbatim}
\NOtes\dotted\islurd0g\qu g\tslur0{'c}\qu c\en
\Notes\dotted\isluru0{'e}\ibl0e{-2}\qb0{edc}\tslur0b\tqb0b\en\bar
\NOtes\dotted\slur{'a}{`f}d1\qu{'a`f}\en
\NOTes\hu g\en
\end{verbatim}\end{quote}

 \section{Modifying font-based slur properties}
Several macros are provided to modify the shape of slurs already initiated.
These macros must be coded right before the slur \ital{termination}. Invoking
any of the macros described in this section will force the slur to be compound.

By default, the midpoint of a font-based slur is three \verb|\internote|s
above or below a line between its ends. This can be changed using the macro
\keyindex{midslur}~$h$ where $h$ is the revised vertical displacement. For
example, \verb|\midslur6| coded right before \verb|\tslur| causes an upper slur to
rise to a maximum height of \verb|6\internote| above the starting position.
For example,

\begin{music}\nostartrule
\startextract
\NOtes\multnoteskip8\isluru0i\ql i\en
\notes\tslur0i\ql i\en
\zendextract
\startextract
\NOtes\multnoteskip8\isluru0i\ql i\en
\notes\midslur7\tslur0i\ql i\en
\zendextract
\startextract
\NOtes\multnoteskip8\isluru2i\ql i\en
\notes\midslur{11}\tslur2i\ql i\en
\zendextract
\end{music}
\noindent was coded as
\begin{quote}\begin{verbatim}
\NOtes\multnoteskip8\isluru0i\ql i\en
\notes\tslur0i\ql i\en
\NOtes\multnoteskip8\isluru0i\ql i\en
\notes\midslur7\tslur0i\ql i\en
\NOtes\multnoteskip8\isluru2i\ql i\en
\notes\midslur{11}\tslur2i\ql i\en
\end{verbatim}\end{quote}
For downward slurs, use negative values; for example,

\begin{music}\nostartrule
\startextract
\NOtes\multnoteskip8\islurd0i\qu i\en
\notes\tslur0i\qu i\en
\zendextract
\startextract
\NOtes\multnoteskip8\islurd0i\qu i\en
\notes\midslur{-7}\tslur0i\qu i\en
\zendextract
\end{music}
\noindent was coded as
\begin{quote}\begin{verbatim}
\NOtes\multnoteskip8\islurd0i\qu i\en
\notes\tslur0i\qu i\en
\NOtes\multnoteskip8\islurd0i\qu i\en
\notes\midslur{-7}\tslur0i\qu i\en
\end{verbatim}\end{quote}

The macro \keyindex{curve}$hij$ allows more precise control over the shape
of a slur or tie. The first parameter $h$
is the vertical deviation and it works exactly like the sole parameter of
\verb|\midslur| described above. The second and
third parameters $i$ and $j$ set the initial and final gradient respectively.
They are defined as the horizontal distance required to attain maximum
vertical deviation. Thus smaller numbers for $i$
and $j$ lead to more extreme gradients.
The default setting is \verb|\curve344|. Hence, coding \verb|\curve322|
doubles the initial and final gradient relative to the default.
As with \verb|\midslur|, \verb|\curve| must be coded {\em immediately} before
the slur termination. The example below illustrates the use of \verb|\curve|.

\begin{music}\nostartrule
\startextract
\NOtes\multnoteskip8\itieu0g\ql g\en
\notes\ttie0\ql g\en
\zendextract
\startextract
\NOtes\multnoteskip8\itieu1g\ql g\en
\notes\curve 322\ttie1\ql g\en
\zendextract
\startextract
\NOtes\multnoteskip8\itieu2g\ql g\en
\notes\curve 111\ttie2\ql g\en
\zendextract
\end{music}
\noindent This was coded as
\begin{quote}\begin{verbatim}
\NOtes\multnoteskip8\itieu0g\ql g\en
\notes\ttie0\ql g\en
\NOtes\multnoteskip8\itieu1g\ql g\en
\notes\curve 322\ttie1\ql g\en
\NOtes\multnoteskip8\itieu2g\ql g\en
\notes\curve 111\ttie2\ql g\en
\end{verbatim}\end{quote}

Two macros are provided to control the behaviour of slurs which extend across
line breaks. Normally, the part of the slur before the line break is
treated as a tie.  This can be changed using \keyindex{breakslur}\enpee,
which sets the termination height of
the broken slur at the line break to pitch $p$, for slur number $n$.

After the line break, the slur is normally resumed at the initial pitch
reference, i.e., the one coded in \verb|\islur|. To change this, the macro
\keyindex{Liftslur}\verb|{|$n$\verb|}{|$h$\verb|}| may be used.
Here $n$ is again the slur reference
number and $h$ is the change in height relative to the initialization height.
This macro is normally used following line breaks, in which case it is best
coded using the \verb|\atnextline| macro. For example, coding
\verb|\def\atnextline{\Liftslur06}| raises the continuation of slur zero
by \verb|6\internote| relative to its initialization height.

The following example illustrates the use of the macros for broken slurs:
\medskip

\begin{minipage}[t]{85mm}
Default, without adjustments:

\begin{music}
\hsize=50mm
\generalmeter{\meterfrac24}%
\parindent 0pt\staffbotmarg0pt%
%
%\interstaff{14\internote}%
%
\nostartrule
\startpiece\bigaccid
\Notes\ibsluru1b\qu b\qu g\en\bar
\Notes\qu{'c!}\tslur1f\qu f\en
\mulooseness1\Endpiece
\end{music}
\begin{verbatim}
\Notes\ibsluru1b\qu b\qu g\en\bar
\Notes\qu{'c!}\tslur1f\qu f\en
\end{verbatim}
\end{minipage}%
\begin{minipage}[t]{75mm}
With~\verb|\Liftslur|~and~\verb|\breakslur|~:\\
\begin{music}
\hsize=50mm
\generalmeter{\meterfrac24}%
\parindent 0pt\staffbotmarg0pt%
%
%\interstaff{14\internote}%
%
\nostartrule
\startpiece\bigaccid
\def\atnextline{\Liftslur17}%
\Notes\ibsluru1b\qu b\qu g\breakslur1g\en\bar
\Notes\qu{'c!}\tslur1f\qu f\en
\mulooseness1\Endpiece
\end{music}
\begin{verbatim}
\def\atnextline{\Liftslur17}%
\Notes\ibsluru1b\qu b\qu g%
\breakslur1g\en\bar
\Notes\qu{'c!}\tslur1f\qu f\en
\end{verbatim}
\end{minipage}\medskip

Occasionally in keyboard works one needs to begin a slur in one
staff but end it in another. This can be done using the macro
\keyindex{invertslur}\onen\ which is best described by
reference to the example
shown below.

% Finding the final pitch is try and error, sorry, no way out.

\begin{music}
\setstaffs1{2}
\setclef1\bass\interstaff{10.5}
\startextract
\NOtes\multnoteskip5\isluru0a\ql a\en
\notes\invertslur0\curve311\tslur0j|\qu d\en
\NOtes\multnoteskip{10}\isluru0a\ql a\en
\notes\invertslur0\curve333\tslur0j|\qu d\en
\zendextract
\end{music}
\noindent This was coded as
\begin{quote}\begin{verbatim}
\NOtes\multnoteskip5\isluru0a\ql a\en
\notes\invertslur0\curve311\tslur0g|\qu d\en
\NOtes\multnoteskip{10}\isluru0a\ql a\en
\notes\invertslur0\curve333\tslur0g|\qu d\en
\end{verbatim}\end{quote}

\noindent Slur inversion as just described takes effect where the slope is zero; therefore it
only works with ascending slurs that were started with \verb|\isluru|, and with
descending slurs started with \verb|\islurd|. Otherwise no horizontal place
can be found and the result is erratic.

A different approach removes this restriction. The idea is to stop the slur at 
the desired inversion point and restart it in the other sense at the
same place. The commands to do this are as follows:
 \begin{itemize}\setlength{\itemsep}{0ex}
 \item \keyindex{Tslurbreak}\enpee~stops slur number $n$ \ital{exactly} at pitch $p$,
not above or below the virtual note head.
 \item \keyindex{Islurubreak}\enpee~restarts an upper slur at the same position,
not above a virtual note head.
 \item \keyindex{Islurdbreak}\enpee~restarts a lower slur at the same position,
not below a virtual note head.
 \end{itemize}

\noindent The vertical position may have to be adjusted to minimize any discontinuity
in the slope. For example, the following pattern

 \begin{music}\nostartrule
\setclef1\treble
\startextract
\NOTes\multnoteskip 3\isluru0a\ql a\en
\NOTes\multnoteskip 3\Tslurbreak0d\Islurdbreak0d\sk\en
\Notes\tslur0h\qu h\en
\NOTes\multnoteskip 3\islurd0a\ql a\en
\NOTes\multnoteskip 3\Tslurbreak0d\Islurubreak0d\sk\en
\Notes\tslur0h\qu h\en
\zendextract
\end{music}
\noindent was coded as
\begin{quote}\begin{verbatim}
\begin{music}\nostartrule
\NOTes\multnoteskip 3\isluru0a\ql a\en
\NOTes\multnoteskip 3\Tslurbreak0d\Islurdbreak0d\sk\en
\Notes\tslur0h\qu h\en
\NOTes\multnoteskip 3\islurd0a\ql a\en
\NOTes\multnoteskip 3\Tslurbreak0d\Islurubreak0d\sk\en
\Notes\tslur0h\qu h\en
\end{music}
\end{verbatim}\end{quote}

\section{Simple slurs}

\ital{Simple} slurs and ties have advantages in some cases: (1) They will
always have the best possible shape, and (2) if \verb|\noteskip| doesn't
change from start to finish, they are easier to code. But they have drawbacks
as well: (1) They are limited in length to 68pt
for slurs and 220pt for
ties, (2) their maximum vertical extent is \verb|8\internote|, and (3) they
may not extend across a line break.
Despite all these limitations, simple slurs are extremely useful in many
applications where the slurs are short and contained within a bar.

Simple slurs must be coded {\em before} the note on which the slur begins.
The primary macro is
\keyindex{slur}\verb|{|$p_1$\verb|}{|$p_2$\verb|}|$sl$
where $p_1$ and $p_2$ are respectively the initial and final pitches,
$s$ is the sense (either ``{\tt u}'' or ``{\tt d}''), and $l$ is the length,
in {\tt noteskip}s.
Thus, thirds slured in pairs can be coded

 %\check

\begin{quote}\begin{verbatim}
\NOtes\slur ced1\qu{ce}\en
\NOtes\slur dfd1\qu{df}\en
\NOtes\slur egd1\qu{eg}\en
\NOtes\slur{'e}cu1\ql{ec}\en
\NOtes\slur{'d}bu1\ql{db}\en
\NOtes\slur{'c}au1\ql{ca}\en
\end{verbatim}\end{quote}
which yields

\begin{music}\nostartrule
\startextract
\NOtes\slur ced1\qu{ce}\en
\NOtes\slur dfd1\qu{df}\en
\NOtes\slur egd1\qu{eg}\en
\NOtes\slur{'e}cu1\ql{ec}\en
\NOtes\slur{'d}bu1\ql{db}\en
\NOtes\slur{'c}au1\ql{ca}\en
\zendextract
\end{music}
%%

There are similar commands to force simple versions of the other
variants of font-based slurs and ties. Simple ties may be set using
\keyindex{tie}\verb|{|$p$\verb|}|$sl$ (only one pitch is needed).
Simple short slurs and short ties can be forced with the macros
\keyindex{sslur}\verb|{|$p_1$\verb|}{|$p_2$\verb|}|$sl$ and
\keyindex{stie}\verb|{|$p$\verb|}|$sl$ respectively. Finally,
simple beam slurs can be forced with
\keyindex{bslur}$s$\verb|{|$p_1$\verb|}{|$p_2$\verb|}|$l$.


\section{Font-based hairpins (crescendos and decrescendos)}\label{sec:crescnd}

There are two categories of font-based hairpins. The first type requires only
one command, \keyindex{crescendo}\verb|{|$\ell$\verb|}| or
\keyindex{decrescendo}\verb|{|$\ell$\verb|}|, where $\ell$ is any \TeX{}
dimension, either a fixed one---for example in points---or a scalable one
expressed either explicitly or implicitly as some number of
\verb|\noteskip|s. These should be used as arguments to
\verb|\zcharnote|, \verb|\zchar|, \verb|\uptext|, \verb|\zmidstaff|, etc., to
post them at the desired altitude. The longest such symbol is $\simeq$ 68 mm.

The second type of font-based hairpin requires two commands, one to start it
and another to end it. The starting macro is \keyindex{icresc}. It has no
arguments. Only one invocation suffices to start any number and combination of
crescendos and diminuendos. The ending macros are \keyindex{tcresc} or
\keyindex{tdecresc}. They should be used as arguments of \verb|\zcharnote| etc,
which will set the altitude. For example,

\begin{music}
\parindent0pt
\generalmeter{\meterfrac{12}8}
\setstaffs1{2}
\startextract
\Notes\cmidstaff\ppp|\ca c\en
\Notes\icresc|\ca{defgh'abcde}\en
\Notes\zmidstaff{\loff\tcresc}\cmidstaff\fff|\ca{'f}\en
\endextract
\end{music}

\noindent which was coded as
\begin{quote}\begin{verbatim}
\Notes\cmidstaff\ppp|\ca c\en
\Notes\icresc|\ca{defgh'abcde}\en
\Notes\zmidstaff{\loff\tcresc}\cmidstaff\fff|\ca{'f}\en
\end{verbatim}\end{quote}
\medskip

\noindent while

\begin{music}
\parindent0pt
\generalmeter{\meterfrac{12}8}
\setstaffs1{2}
\startextract
\Notes\cmidstaff\ppp|\ca c\en
\Notes\icresc|\ca{defgh'abcde}\en
\Notes\zcharnote N{\tcresc}\cmidstaff\fff|\zcharnote q{\tcresc}\ca{'f}\en
\endextract
\end{music}

\noindent was coded as
\begin{quote}\begin{verbatim}
\Notes\cmidstaff\ppp|\ca c\en
\Notes\icresc|\ca{defgh'abcde}\en
\Notes\zcharnote N{\tcresc}\cmidstaff\fff|\zcharnote q{\tcresc}\ca{'f}\en
\end{verbatim}\end{quote}

 \section{Limitations of font-based slurs}
  The change in altitude between slur initiation and slur termination is limited to
\verb|16\Internote|. Thus unexpected vertical gaps can appear, as seen in

 \begin{music}\nostartrule
\instrumentnumber{1}
\generalmeter{\meterfrac34}
\startextract\NOTes\multnoteskip3\isluru0c\ql c\tslur0j\ql j\en
  \bar\NOTes\multnoteskip3\isluru0c\ql c\tslur0n\ql n\en\zendextract

\startextract\NOTes\multnoteskip3\isluru0c\ql c\tslur0s\ql s\en
  \bar\NOTes\multnoteskip3\isluru0c\ql c\tslur0z\ql z\en\zendextract
 \end{music}
 \noindent whose coding was
 \begin{quote}\begin{verbatim}
\NOTes\multnoteskip3\isluru0c\ql c\tslur0j\ql j\en
  \bar\NOTes\multnoteskip3\isluru0c\ql c\tslur0n\ql n\en
\NOTes\multnoteskip3\isluru0c\ql c\tslur0s\ql s\en
  \bar\NOTes\multnoteskip3\isluru0c\ql c\tslur0z\ql z\en
 \end{music}
 \end{verbatim}\end{quote}

Furthermore, if the slope becomes too steep, even worse results can occur, such as

\medskip
 \begin{music}\nostartrule
\instrumentnumber{1}
\generalmeter{\meterfrac34}
\startextract\NOTes\isluru0c\ql c\en\notes\tslur0j\ql j\en
  \bar\NOTes\isluru0c\ql c\en\notes\tslur0n\ql n\en
 \NOTes\isluru0c\ql c\en\notes\tslur0s\ql s\en
 \bar\NOTes\isluru0c\ql c\en\notes\tslur0z\ql z\en
 \zendextract
 \end{music}

Another limitation of font-based slurs crops up when one attempts to
generate bitmapped versions for high-resolution printers: the characters
for long ties can exceed \MF's maximum capacity.
